---
layout: post
title: "Spatial Intro 03: Data About Data -- Intro to Metadata File Formats and Structure"
date:   2015-10-27
authors: [Leah A. Wasser, Megan A. Jones]
contributors: [ ]
dateCreated: 2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [raster, rgdal, eml, devtools]
category: [self-paced-tutorial] 
tags: [R, spatial-data-gis, metadata]
mainTag: spatial-data-management-series
workshopSeries: [spatial-data-management-series]
description: "This tutorial covers what metadata are, and why we need to work with
metadata. It discusses metadata in the context of spatio-temporal data. It 
covers the 3 most common metadata formats: text file format, 
web page format and Ecological Metadata Language (EML)."
code1: 03-metadata-formats-and-files.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: 
permalink: R/metadata-file-formats-structures
comments: true
---

{% include _toc.html %}

## About
This tutorial covers what metadata are, and why we need to work with
metadata. It discusses metadata in the context of spatio-temporal data. It 
covers the 3 most common metadata formats: text file format, 
web page format and Ecological Metadata Language (EML).

**R Skill Level:** Introduction - You are just learning about metadata.

<div id="objectives" markdown="1">

# Goals / Objectives

After completing this activity, you will:

* Understand that metadata come in many formats and must be maintained with the
data. 
* Understand that there is necessary metadata associated with and/or embedded in
the data.
* Understand that collecting data requires/includes collecting the metadata.
* Be able to create metadata files for different data types. 

## Things You’ll Need To Complete This Lesson
To complete this lesson you will need the most current version of R, and 
preferably, RStudio loaded on your computer.

### Install R Packages

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`

**OPTIONAL**

* **devtools** `install.packages("devtools")` 
* **eml** `install_github("ropensci/EML", build=FALSE, dependencies=c("DEPENDS", "IMPORTS"))`


* [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

### Download Data

****

{% include/_greyBox-wd-rscript.html %}

**Spatial-Temporal Data & Data Management Lesson Series:** This lesson is part
of a lesson series introducing
[spatial data and data management in `R` ]({{ site.baseurl }}tutorial/URL).
It is also part of a larger 
[spatio-temporal Data Carpentry Workshop ]({{ site.baseurl }}workshops/spatio-temporal-workshop)
that includes working with  
[raster data in `R` ]({{ site.baseurl }}tutorial/spatial-raster-series),
[vector data in `R` ]({{ site.baseurl }}tutorial/spatial-vector-series)
and  
[tabular time series in `R` ]({{ site.baseurl }}tutorial/tabular-time-series).

****

### Additional Resources

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">

</div>


```{r elevation-map, include=TRUE, results="hide", echo=FALSE}
#render DSM for lesson content background
DSM_HARV <- raster("NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

# code output here - DEM rendered on the screen
plot(DSM_HARV,
      main="A Dataset You Are Given\n What metric does it represent?\nHow Was It Processed??")

```


<div id="challenge" markdown="1">
## Challenge: What Do I Know About My Data?

The figure above was create from a that you are given called: `HARV_dsmCrop.tif`.  
What other information would you like to know about the data used to create the 
above map before you would feel comfortable using this data to address a 
research question? 

</div>

```{r challenge-1, echo=FALSE}

# the idea is that we have no idea what this raster represents (what metric),
# what the units are, what the scale represents. When it was collected. 
# this should engage the participants in a discussion of how do we share / document
# data efficiently for sharing.

```

## Why Do We Need Metadata?

Looking at the map above, we are missing information needed to begin working with
the data effectively including:

**Spatial Information**

* **Spatial Extent:** What area does this dataset cover? 
* **Coordinate reference system:** what spatial projection / coordinate reference
system are used to store the data? Will it line up with other data? 
* **Resolution:** The data appears to be in **raster** format. This means it is
composed of pixels. What area on the ground does each pixel cover - ie What is 
its spatial resolution?

**Data Collection / Processing Methods**

* **How was this data generated?** Is this an output from a model, is it an image
from a remote sensing instrument such as a satellite (e.g. landsat) or collected
from an airplane? How were the data colected?
* **Units:** We can see a scale bar of values to the right of the data however, 
what units / metric does this represent? Temperature? Elevation? Precipitation?
* **How were the data processed?**

**Contact Information**
* Who created this data? Who do we contact if we have questions?

When we are given a dataset, or when we download it online, we do not know anything
about it without proper documentation. This documentation is called **metadata** - 
data about the data.

## What is Metadata?
Metadata is structured information that describes a dataset. It includes a suite of
information about the data including:

* Contact information
* Spatial Attributes including: extent, coordinate reference system, resolution
* Data collection & processing methods
* and more.

Without sufficient documentation, it is difficult for us to work with 
external data - data that we did not collect ourselves.

### Why Are Metadata Needed?
We need metadata to not only work with external data, but also when it's embedded
in a file or in provided in a machine readable format, we can access it directly
in tools like `R` or `Python`. We will talk about different metadata formats,
next.

## Metadata Formats

Metadata come in different formats. We will discuss three of those in this tutorial:

* **Structured Embedded Metadata:** Some file formats (e.g. `Geotiff` and `HDF5`)
 supported embedded metadata which you can access from a tool like `R` or `Python`
 directly when you import the data.
* **Structured Metadata Files:** Structured metadata files such as the Ecological 
Metadata Language (EML) are stored in a machine readable format which means they
can be accessed using a tool like `R` or `Python`. There are different file formats
and standards but it's important to understand that standards will vary.
* **UnStructured Metadata Files:** These include text files, web pages and other
documentation that does not follow a particular standard or format, but documents
key attributes required to work with the data.


<i class="fa fa-star"></i> **Data Tip:** When you find metadata for a dataset 
that you are working with, **DOWNLOAD AND SAVE IT** immediately to a directory 
on your computer where you saved the data. It is also a good idea to document
the URL where you found the metadata and the data in a "readme" text file!
{: .notice}


## Embedded Metadata - Geotiff

If we want to automate workflows, it's ideal when key metadata required to process
our data are embedded directly in our data files. The **geotiff** (`fileName.tif`) 
is one common spatio-temporal format that can store **metadata** directly in the
`.tif` file itself.

## What is a GeoTIFF??

A `geotiff` file stores metadata or attributes about the file as embedded `tif tags`. 
A GeoTIFF is a standard `.tif` image format with additional spatial (georeferencing)
information embedded in the file as tags. These tags can include the following raster metadata:

1. A Coordinate Reference System (`CRS`)
2. Spatial Extent (`extent`)
3. Values for when no data is provided (`NoData` Value)
4. The `resolution` of the data

<i class="fa fa-star"></i> **Data Tip:**  Your camera uses embedded tags to store
information about pictures that you take including the camera make and model, 
and the time the image was taken!
{: .notice }

More about the  `.tif` format:

* <a href="https://en.wikipedia.org/wiki/GeoTIFF" target="_blank"> GeoTIFF on Wikipedia</a>
* <a href="https://trac.osgeo.org/geotiff/" target="_blank"> OSGEO TIFF documentation</a>

The `Raster` package in `R` allows us to directly access tif tags programatically.
We can quickly view the spatial `extent`, `coordinate reference system` and `resolution` 
of the data. 

```{r load-libraries }
# load libraries
library(raster)
library(rgdal)

# set working directory to ensure R can find the file we wish to import
# setwd("working-dir-path-here")

# render DSM for lesson content background
DSM_HARV <- raster("NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

# view Coordinate Reference System (note this often contains horizontal units!)
crs(DSM_HARV)

# assign crs to an object (class) to use for reprojection and other tasks
myCRS <- crs(DSM_HARV)
myCRS

# view spatial extent
extent(DSM_HARV)

# view spatial resolution
res(DSM_HARV)

DSM_HARV
```


We can use embedded metadata to programtically perform processing tasks 
on our data including reprojections, cropping and more.

We will work with eembedded metadata in both the [vector data]( http://www.neondataskills.org/tutorial-series/vector-data-series/)  and 
[raster data]( http://www.neondataskills.org/tutorial-series/raster-data-series/) 
series!

### Hierarchical Data Formats (HDF5)

HDF5 is another file type that supports embedded metadata format. Check out the 
HDF5 tutorial series to learn more about 
[how HDF5 stores metadata](http://www.neondataskills.org/tutorial-series/intro-hdf5-r-series/).


## Structured Metadata - EML

Metadata can also be stored, separately, in a separate document. This document
may be: 

1. **Carefully structured** So that we can access metadata contents programatically
(e.g. EML)
2. **Written out in text or web page format** This is less ideal as it means we have
to manually extract key metadata attributes to use in our code but it is often useful
if we are reading about data collection and processing methods which may be 
length and detailed to support reproducible workflows.

We will briefly look at both structured and unstructured (non-embedded) metadata 
formats, next.


## Structured Metadata - Introduction to EML

The Ecological Metadata Language (EML) is a data specification developed specifically
to document ecological data. An EML file is created using a `XML` based format.
This means that content is embedded within hierarchical tags. For example
the title of a dataset might be embedded in a `<title>` tag as follows:

`<title>Fisher Meteorological Station at Harvard Forest since 2001</title>`

Similarly, the creator of a dataset is also be found in a hierarchical tag
structure.


    <creator>
      <individualName>
        <givenName>Emery</givenName>
        <surName>Boose</surName>
      </individualName>
    </creator>
    

The `EML` package for `R` is designed to open, read and create `EML`
formatted metadata. In this tutorial, we will overview demonstrate how we can
use EML in an automated workflow. NOTE: To save time, we will not explicitly 
teach the `EML` package given it is still being developed. But we will provide
an example of how you can access `EML` programmatically below.

## EML Terminology

Let's first discuss some basic `EML` terminology. In the 
context of `EML`, a file documents a `dataset`. This `dataset` may consist of one
or more files that are documented in the `EML` document. In the case of our 
tabular meteorology data, the structure of our `EML` file includes:

1. The **dataset**. A dataset may contain
one or more data tables associated with it that may contain different types of related
information. For this Harvard meteorological data, the data tables contain tower
measurements including precipitation and temperature, that are aggregated
at various time intervales (15 minute, daily, etc) and that date back to 2001.
2. The **data tables**. Data tables refer to the actual data that make up the dataset. 
For the Harvard Data, each data table contains a suite of meterological metrics 
including precipiation and temperature (and associated quality flags), that are 
aggregated at a particular time interval. (e.g. one data table contains monthly
average data, another contains 15 minute averaged data, etc)

## Work With EML in R 

To begin, we will load the `EML` package directly from ropensci's Git repository.


```{r install-EML-package, results="hide", warning=FALSE }
# install R EML tools
# library("devtools")
# install_github("ropensci/EML", build=FALSE, dependencies=c("DEPENDS", "IMPORTS"))

# load ROpenSci EML package
library("EML")

# load ggmap for mapping
library(ggmap)


# data location
# http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001
# table 4 http://harvardforest.fas.harvard.edu/data/p00/hf001/hf001-04-monthly-m.csv
```

Next, we will read in the LTER `EML` file - directly from the online URL using
`eml_read`. This file documents multiple data products that can be downloaded.
<a href="http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001" target="_blank">
Check out the Harvard Forest Data Archive Page for Fisher Meteorological Station
for more on this dataset and to download the archive files directly.</a>

Note that because this eml file is large, it takes a few seconds for the file to 
load.


```{r read-eml }
# import EML from Harvard Forest Met Data
eml_HARV <- eml_read("http://harvardforest.fas.harvard.edu/data/eml/hf001.xml")

# view size of object
object.size(eml_HARV)

# view the object class
class(eml_HARV)
```

The `eml_read` function creates an `EML` class object. This object can be accessed
using `slots` in R (`@`) rather than a typical subset `[]` approach.

## Explore Metadata Attributes

We can begin to explore the contents of our EML file and associated data that it
describes. Let's start at the dataset level. We can use `eml_get` to view the 
contact information for the dataset, the keywords and it's associated temporal
and spatial (if relevant) coverage.


```{r view-eml-content }
# view the contact name listed in the file
# this works well!
eml_get(eml_HARV,"contact")

# grab all keywords in the file
eml_get(eml_HARV,"keywords")

# figure out the extent & temporal coverage of the data
eml_get(eml_HARV,"coverage")

```


## Identify & Map Data Location

Looking at the coverage for our data, there is only one unique x and y value. This 
suggests that our data were collected at (x, y) one point location. We know this is a 
tower so a point location makes sense. Let's grab the x and y coordinates and 
create a quick context map. We will use `ggmap` to create our map.

**NOTE: if this were a rectangular extent we'd want the bounding BOX. this is important
if the data are for example, raster format, in HDF5 or something. we need the extent
to properly geolocate and process the data.**

<a href="https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf" target="_blank">Learn More: A nice cheatsheet for GGMAP created by NCEAS</a>

```{r map-location, warning=FALSE, message=FALSE}
# grab x coordinate
XCoord <- eml_HARV@dataset@coverage@geographicCoverage@boundingCoordinates@westBoundingCoordinate
# grab y coordinate
YCoord <- eml_HARV@dataset@coverage@geographicCoverage@boundingCoordinates@northBoundingCoordinate


# map <- get_map(location='Harvard', maptype = "terrain")
map <- get_map(location='massachusetts', maptype = "toner", zoom =8)

ggmap(map, extent=TRUE) +
  geom_point(aes(x=XCoord,y=YCoord), 
             color="darkred", size=6, pch=18)

```

The above example, demonstated how we can extract information from an `EML` document
and use it programatically in `R`! This is just the beginning of what we can do!


## Unstructured Metadata - Web pages & Text Files

## Metadata Stored on a Web Page

Some metadata are stored in a less or non-structured format such as a web page,
or a text file. For example, if we  <a href="http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001" target="_blank">visit the Harvard Forest Fisher Tower Webpage</a>, we find
that metadata is stored there for some of the data that we use in the 
[tabular time tutorial series](http://www.neondataskills.org/tutorial-series/tabular-time-series/).

If we visit the link above, we see a list of data available for Harvard Forest. 

Scroll down to the **Overview** section. Take note of the information provided 
in that section and answer the following questions in the challenge below:

<div id="challenge" markdown="1">
## Challenge - Explore Metadata

Explore the metadata stored on the Harvard Forest LTER web page. Answer the 
following questions. 

1. What is the time span of the data available for this site? 
2. You have some questions about these data. Who is the lead investigator / who 
do you contact for more information? And how do you contact them? 
3. Where is this field site located? How is the site location information stored
in the metadata? Is there any information that may be useful to you viewing the location?
(HINT: what if you were not familiar with Harvard as a site / from another country, 
etc?)
4. Field Site Information: What is the elevation for the site? What is the dominant
vegetation cover for the site? HINT: is dominant vegetation easy to find in the
metadata?
5. How is snow dealt with in the precipitation data?
6. Are there some metadata attributes that might be useful to access in a script
in `R` or `Python` rather than viewed on a web page?

HINT: Can you answer all of the questions above from the information provided
on this website? Is there information that you might prefer to find on that page?
</div>

```{r metadata-challenge-1, echo=FALSE}
# Metadata Notes from hf001_10-15-m_Metadata.txt
# 1. 2001-2015
# 2. Emery Boos - located at the top of the document, email is available
# 3. a lat long is available in the metadata at the top, we see the location described
# as Prospect Hill Tract (Harvard Forest). 
# 4. 342 m elevation, the veg type is not clear in the metadata
# 5. Found in the methods: Delayed melting of snow and ice (caused by problems with rain gage heater or heavy precipitation) is noted in log - daily values are corrected if necessary but 15-minute values are not. The gage may underestimate actual precipitation under windy or cold conditions.
# 6. this could be a discussion. things like units, time zone, etc are all useful
# if accessed programmatically

```

## Unstructured vs Structured Metadata

Metadata are particularly important when we are working with data that we did not
collect ourselves. When you download, or gain access to data from a colleague, 
or other data provider, be sure to first find and review the metadata. Then make
sure you save the metadata in a directory that is closely associated with wherever
you save the data itself! Structured metadata formats are ideal if you can find
them because they are most often:

* In a standard, documented format that others use.
* Are machine readable which means you can use them in scripts and algorithms.




## Create Metadata For Your Data

When you are creating data that you want to share with others, it is critical
to create your own metadata. While this is beyond the scope of this tutorial,
There are many resources available to support metadata documentation including:

* Morpho
* others...

# Be sure to link to the metadata support pages and tools
# a good challenge for this would be to find a few datasets and have them explore the 
site and find the metadata for them -- NLCD is a good example.


## ATBD vs Metadata

Algorithm Theoretical Basis Document (ATBD) is a document which describes the 
collection and processing methods associated with a data product. While this topic
is beyond the scope of this tutorial, it is important to note that often times
some important information about your data, is documented in other documents
such as ATBDs. <a href="http://www.neonscience.org" target="_blank">NEON</a> provides
ATBDs for all of it's data products.


